---
title: "MIRAGEdata"
purpose: To load, merge, explore, and plot lipidomics data and look for correlations with cardiometabolic endpoints the MIRAGE dataset.
output:
html_notebook: default
author: Ole Emil Andersen
date created: March 2024
editor_options: 
  chunk_output_type: console
---

# Packages

```{r}
library(readxl) # reads Excel files
library(dplyr) # to manipulate data
library(tidyr) # facilitates tidy data transformation.
library(purrr) # enhances programming capabilities
library(janitor) # to clean column names
library(ggplot2) # for plots
library(lme4) # for mixed models
library(emmeans) # for p-values and confidence intervals
```

# Quantification limits

Values are inserted from the lipidomics Excel file, "2022Mar Goodpaster Results.xlsx"

```{r}
detection_limits <- data.frame(
  lipidClass = c("SPH", "AC", "SPM", "PG", "LPG", "PI", "LPI", "PS", "LPS", "PA", "PE", "LPE", "PC", "LPC", "1,3-DAG", "1,2-DAG", "Cer", "dHCer", "GluCer", "LacCer", "GalCer", "TAG"),
  LLOQ = c(0.08, 0.16, 1.2, 0.6, 0.03, 2.5, 1, 2.5, 0.05, 0.1, 2.5, 1, 7, 3, 0.34, 0.68, 0.25, 0.05, 0.1, 0.2, 0.1, 10),
  ULOQ = c(320, 640, 4800, 2400, 100, 10000, 4000, 10000, 200, 400, 10000, 4000, 28000, 12000, 1360, 2720, 1000, 200, 400, 800, 400, 40000)
)
```

# Load data

**Load lipidomics** data into long format and clean the data by replacing \<0 with 0 and N/A with NA.

```{r}
# Define the path to the Excel file
file_path <- "/Users/au309166/Library/CloudStorage/OneDrive-Aarhusuniversitet/Forskning/Projects/LipidomicsMIRAGE/data-raw/2022Mar Goodpaster Results With Name Spaces.xlsx"

# List sheet names to load from excel file
sheet_names <- c("SPM SPH AC", "PG", "PI", "PS", "PA", "PE", "PC", "1,3-DAGs", "1,2-DAGs", "Cer dHCer GluCer LacCer GalCer", "TAG")

load_sheets_in_long_format <- function(file_path, sheet_names) {
  map_df(sheet_names, ~ {
    # Read the measurement names from the first row
    measurement_names <- read_excel(file_path, sheet = .x, n_max = 1) %>%
      colnames()

    # Read the data, skipping the first two rows
    data <- read_excel(file_path, sheet = .x, skip = 2, col_names = FALSE)

    # Limit the data to match the number of columns in the column names
    num_cols_in_names <- length(measurement_names)
    data_trimmed <- data[, 1:num_cols_in_names]

    # Assign column names to the data
    colnames(data_trimmed) <- c("Sample Name", measurement_names[-1])

    # Adjusting the replacement for "< 0" with 0 (as numeric) and "N/A" with NA
    data_cleaned <- data_trimmed %>%
      mutate(across(everything(), ~ case_when(
        .x == "< 0" ~ "0", # Replace "< 0" with "0"
        .x == "N/A" ~ NA_character_, # Replace "N/A" with NA
        TRUE ~ as.character(.x) # Keep other values as-is, converting to character to ensure type consistency
      ))) %>%
      type.convert(as.is = TRUE) # Convert back to original type where possible

    # Rename "Sample Name" to "sampleNumber"
    colnames(data_cleaned)[1] <- "sampleNumber"

    # Transform to long format
    data_long <- pivot_longer(data_cleaned, -sampleNumber, names_to = "measurement", values_to = "pmol")

    # Split the measurement into class and structure based on space
    data_long <- data_long %>%
      separate(measurement, into = c("lipidComposition", "lipidClass"), sep = " ", extra = "merge")

    return(data_long)
  }, .id = "sheet")
}

# Load all specified sheets into a single data frame in long format
lipidomicsDF <- load_sheets_in_long_format(file_path, sheet_names)

# Remove sheet column
lipidomicsDF <- select(lipidomicsDF, -sheet)
```

**Load identifiers and sample weights** and merge with the data frame and split the pid into pre/post and id

```{r}
# path to file
filePathIdentifiers <- "/Users/au309166/Library/CloudStorage/OneDrive-Aarhusuniversitet/Forskning/Projects/LipidomicsMIRAGE/data-raw/Lipidomics Key and Weights Goodpaster KZB_identifiable.xlsx"

# load file
identifiersDF <- read_excel(filePathIdentifiers)

# clean names to make easier to work with
identifiersDF <- clean_names(identifiersDF)

# Make tube_id character
identifiersDF$tube_id <- as.character(identifiersDF$tube_id)

# Combine identifiers with data
lipidomicsDF <- left_join(lipidomicsDF, identifiersDF, by = c("sampleNumber" = "tube_id"))

# Split pid column into an id and a time column
lipidomicsDF <- separate(lipidomicsDF, pid, into = c("id", "time"), sep = " ", extra = "merge")

# Make time, lipidClass, id and group categorical variables
lipidomicsDF$time <- factor(lipidomicsDF$time, levels = c("Pre", "Post")) # sort by pre and post
lipidomicsDF$lipidClass <- factor(lipidomicsDF$lipidClass)
lipidomicsDF$id <- factor(lipidomicsDF$id)
lipidomicsDF$group <- factor(lipidomicsDF$group)
lipidomicsDF$lipidComposition <- factor(lipidomicsDF$lipidComposition)
```

**Load other endpoints** from "FINAL MIRAGE DATASET.xlsx".

!!! DUPLICATE IDS: "MG023" and "MG019" both are duplicates that appear with same group as in id from Pittsburg and TRI. Assumption: All samples are from TRI if not specified otherwise. Therefore, the Excel file "FINAL MIRAGE DATASET.xlsx", sheet "Not Imputed-Clean" was changed to make the merge work. All changes are marked with red text.

```{r}
# path to file
file_path_endpoints <- "/Users/au309166/Library/CloudStorage/OneDrive-Aarhusuniversitet/Forskning/Projects/LipidomicsMIRAGE/data-raw/FINAL MIRAGE DATASET.xlsx"

# load file
mirage_data <- read_excel(file_path_endpoints, sheet = "Not Imputed-Clean")

# clean names
mirage_data <- clean_names(mirage_data)

# separate pid into id and site
mirage_data <- separate(mirage_data, pid, into = c("id", "site"), sep = "-", extra = "merge")

# Remove redundant columns
mirage_data <- select(mirage_data, -study_center)

# This step not needed? Merge data with lipidomics data
# lipidomicsDF <- left_join(lipidomicsDF, mirage_data, by = 'id')
```

# Within thresholds

Make column to denote if measurement is within (0), below LLOQ (1) or above ULOQ (2)

```{r}
# Initialize the "within_threshold" column in lipidomicsDF
lipidomicsDF$within_threshold <- NA

# Iterate over lipidomicsDF and assign "within_threshold" values
for (i in 1:nrow(lipidomicsDF)) {
  # Extract the current row's lipidClass
  current_lipidClass <- lipidomicsDF$lipidClass[i]

  # Lookup the corresponding LLOQ and ULOQ values in detection_limits
  current_LLOQ <- detection_limits$LLOQ[detection_limits$lipidClass == current_lipidClass]
  current_ULOQ <- detection_limits$ULOQ[detection_limits$lipidClass == current_lipidClass]

  # Check for NA values in pmol or in the LLOQ/ULOQ lookup
  if (is.na(lipidomicsDF$pmol[i]) | length(current_LLOQ) == 0 | length(current_ULOQ) == 0) {
    lipidomicsDF$within_threshold[i] <- NA # Assign NA if there's an NA value in pmol or LLOQ/ULOQ cannot be found
  } else if (lipidomicsDF$pmol[i] < current_LLOQ) {
    lipidomicsDF$within_threshold[i] <- 1 # Below LLOQ
  } else if (lipidomicsDF$pmol[i] > current_ULOQ) {
    lipidomicsDF$within_threshold[i] <- 2 # Above ULOQ
  } else {
    lipidomicsDF$within_threshold[i] <- 0 # Within the threshold
  }
}
```

# Normalization and transformation

Weight normalization

```{r}
lipidomicsDF$pmol_mg <- lipidomicsDF$pmol / lipidomicsDF$equiv_weight_extracted
```

Log transform the weight normalized lipidomics data. 1 is added to all values to avoid zores.

```{r}
# Adding a small constant (e.g., 1) to avoid log(0)
lipidomicsDF <- lipidomicsDF %>%
  mutate(pmol_mg_log = log(pmol_mg + 1))
```

# Sort data

Sort data for future plotting

```{r}
# Sort the dataframe by 'id' and 'time'
lipidomicsDF <- lipidomicsDF %>%
  arrange(id, time)
```

# Scatter plot

Scatter plots with insulin sensitivity as colormap. Datapoints above ULOQ and below LLOQ are marked with triangles for.

```{r}
# Create a named vector mapping IDs to continuous values for coloring
color_map <- setNames(mirage_data$gdr_pre_rd_ss2_mg_kg_ffm_min, mirage_data$id)

plot_lipid <- function(lipidomicsDF, lipid_composition, lipid_class, color_map, detection_limits) {
  filtered_data <- filter(lipidomicsDF, lipidClass == lipid_class & lipidComposition == lipid_composition & !is.na(time))

  # Find the min and max of the y-axis data
  y_min <- min(filtered_data$pmol_mg, na.rm = TRUE)
  y_max <- max(filtered_data$pmol_mg, na.rm = TRUE)

  # Optionally, you can add a buffer to these limits
  buffer <- (y_max - y_min) * 0.05 # 5% buffer, adjust as necessary
  y_min <- y_min - buffer
  y_max <- y_max + buffer

  # Prepare the plot with the existing setup
  plot <- ggplot(
    filtered_data,
    aes(
      x = factor(interaction(group, time, sep = " "),
        levels = c("HED Pre", "HED Post", "DIWL Pre", "DIWL Post", "WLEX Pre", "WLEX Post")
      ),
      y = pmol_mg, group = interaction(id, group)
    )
  ) +
    geom_line(aes(group = interaction(id, group)), color = "darkgrey", position = position_nudge(x = 0.1)) + # Nudge lines to the right
    geom_point(aes(color = color_map[id]), alpha = 0.5, position = position_nudge(x = 0.1)) + # Nudge points to the right
    geom_boxplot(aes(group = interaction(group, time)), width = 0.3, alpha = 0.2, outlier.shape = NA, position = position_dodge(width = 0.75), color = "black") +
    geom_point(
      data = subset(filtered_data, within_threshold == 2),
      aes(
        x = factor(interaction(group, time, sep = " "),
          levels = c("HED Pre", "HED Post", "DIWL Pre", "DIWL Post", "WLEX Pre", "WLEX Post")
        ),
        y = pmol_mg
      ),
      shape = 17, # Triangle up
      color = "black",
      alpha = 0.3, # make translucent
      position = position_nudge(x = 0.1)
    ) +
    geom_point(
      data = subset(filtered_data, within_threshold == 1),
      aes(
        x = factor(interaction(group, time, sep = " "),
          levels = c("HED Pre", "HED Post", "DIWL Pre", "DIWL Post", "WLEX Pre", "WLEX Post")
        ),
        y = pmol_mg
      ),
      shape = 25, # Triangle down
      color = "black",
      alpha = 0.3, # make translucent
      position = position_nudge(x = 0.1)
    ) +
    scale_color_gradient(low = "blue", high = "red", name = "Glucose disposal rate,\n(mg/kgFFM/min)") +
    labs(title = paste(lipid_composition, lipid_class), y = "pmol/mg", x = "") +
    theme_minimal() +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      legend.title.align = 0.5,
      plot.title = element_text(hjust = 0.5),
      legend.text = element_text(size = 6),
      legend.title = element_text(size = 6),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )

  # Filter the detection_limits for the current lipid_class
  limits <- detection_limits %>%
    filter(lipidClass == lipid_class)

  # Set the y-axis limits based on the data range
  plot <- plot + coord_cartesian(ylim = c(y_min, y_max))

  # Define the filename, making it OneDrive compatible
  filename <- paste0(
    "/Users/au309166/Library/CloudStorage/OneDrive-Aarhusuniversitet/Forskning/Projects/LipidomicsMIRAGE/outputs/scatterplots_exploration/",
    gsub("/", "_", gsub(":", ",", paste(lipid_class, lipid_composition, sep = " "))),
    ".jpg"
  )

  # Save the plot as a high-resolution JPEG
  ggsave(filename, plot = plot, device = "jpeg", width = 8, height = 6, dpi = 300)

  cat("Plot saved as:", filename, "\n")
}

# Extract unique lipid classes
unique_lipid_classes <- unique(lipidomicsDF$lipidClass)

# Loop through each lipid class
for (lipid_class in unique_lipid_classes) {
  # Filter for unique lipid compositions within the current lipid class
  lipid_compositions_in_class <- unique(lipidomicsDF$lipidComposition[lipidomicsDF$lipidClass == lipid_class])

  # Loop through each lipid composition within the current lipid class
  for (lipid_composition in lipid_compositions_in_class) {
    # Call the plotting function for each combination of lipid_class and lipid_composition
    plot_lipid(lipidomicsDF, lipid_composition, lipid_class, color_map, detection_limits)
  }
}
```

# Make QQ plots

Raw, normalized lipidomics data

```{r}
output_dir <- "/Users/au309166/Library/CloudStorage/OneDrive-Aarhusuniversitet/Forskning/Projects/LipidomicsMIRAGE/outputs/qq-plots/raw"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

unique_combinations <- lipidomicsDF %>%
  distinct(lipidClass, lipidComposition) %>%
  arrange(lipidClass, lipidComposition)

groups <- c("HED", "DIWL", "WLEX")
times <- c("Pre", "Post")

for (i in seq_len(nrow(unique_combinations))) {
  current_combination <- unique_combinations[i, ]

  sanitized_lipidClass <- gsub("[/:]", "_", current_combination$lipidClass)
  sanitized_lipidComposition <- gsub("[/:]", "_", current_combination$lipidComposition)
  file_name <- paste(sanitized_lipidClass, sanitized_lipidComposition, "qqplots.jpeg", sep = " ")
  file_path <- file.path(output_dir, file_name)

  # Open JPEG device
  jpeg(file_path, width = 800, height = 600)

  # Setup plot layout
  par(mfrow = c(3, 2), mar = c(4, 4, 2, 1))

  for (group in groups) {
    for (time in times) {
      data_to_plot <- lipidomicsDF %>%
        filter(
          lipidClass == current_combination$lipidClass,
          lipidComposition == current_combination$lipidComposition,
          group == !!group,
          time == !!time
        )

      # Plot only if there's data
      if (nrow(data_to_plot) > 0 && !all(is.na(data_to_plot$pmol_mg))) {
        qqnorm(data_to_plot$pmol_mg, main = paste(current_combination$lipidClass, current_combination$lipidComposition, group, time))
        qqline(data_to_plot$pmol_mg, col = "red")
      } else {
        plot.new()
        title(main = paste("No data for", current_combination$lipidClass, current_combination$lipidComposition, group, time))
      }
    }
  }

  # Close the plotting device
  dev.off()
}

cat("All plots have been saved to", output_dir, "\n")
```

Log transformed lipidomics data

```{r}
output_dir <- "/Users/au309166/Library/CloudStorage/OneDrive-Aarhusuniversitet/Forskning/Projects/LipidomicsMIRAGE/outputs/qq-plots/log transformed"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

unique_combinations <- lipidomicsDF %>%
  distinct(lipidClass, lipidComposition) %>%
  arrange(lipidClass, lipidComposition)

groups <- c("HED", "DIWL", "WLEX")
times <- c("Pre", "Post")

for (i in seq_len(nrow(unique_combinations))) {
  current_combination <- unique_combinations[i, ]

  sanitized_lipidClass <- gsub("[/:]", "_", current_combination$lipidClass)
  sanitized_lipidComposition <- gsub("[/:]", "_", current_combination$lipidComposition)
  file_name <- paste(sanitized_lipidClass, sanitized_lipidComposition, "Log qqplots.jpeg", sep = " ")
  file_path <- file.path(output_dir, file_name)

  # Open JPEG device
  jpeg(file_path, width = 800, height = 600)

  # Setup plot layout
  par(mfrow = c(3, 2), mar = c(4, 4, 2, 1))

  for (group in groups) {
    for (time in times) {
      data_to_plot <- lipidomicsDF %>%
        filter(
          lipidClass == current_combination$lipidClass,
          lipidComposition == current_combination$lipidComposition,
          group == !!group,
          time == !!time
        )

      # Plot only if there's data
      if (nrow(data_to_plot) > 0 && !all(is.na(data_to_plot$pmol_mg))) {
        qqnorm(data_to_plot$pmol_mg_log, main = paste(current_combination$lipidClass, current_combination$lipidComposition, group, time))
        qqline(data_to_plot$pmol_mg_log, col = "red")
      } else {
        plot.new()
        title(main = paste("No data for", current_combination$lipidClass, current_combination$lipidComposition, group, time))
      }
    }
  }

  # Close the plotting device
  dev.off()
}

cat("All plots have been saved to", output_dir, "\n")
```

# Mixed models

Insulin sensitivity. Original data was analyzed by one-way ANOVA. So this would be to double check results

```{r}
# Convert from wide to long format and select specific columns
mirage_data_long <- mirage_data %>%
  pivot_longer(cols = c("gdr_pre_rd_ss2_mg_kg_ffm_min", "gdr_post_rd_ss2_mg_kg_ffm_min"),
               names_to = "time",
               values_to = "glucose_uptake") %>%
  mutate(time = recode(time,
                       "gdr_pre_rd_ss2_mg_kg_ffm_min" = "Pre",
                       "gdr_post_rd_ss2_mg_kg_ffm_min" = "Post")) %>%
  select(id, time, glucose_uptake, group)

# Set reference groups before analysis
mirage_data_long$group <- factor(mirage_data_long$group, levels = c("HED", "DIWL", "WLEX"))
mirage_data_long$time <- factor(mirage_data_long$time, levels = c("Pre", "Post"))





# Simple boxplot
boxplot(glucose_uptake ~ time * group, data = mirage_data_long)

# 'insulin_sensitivity' is the dependent variable
# Fixed effects: time, group, and their interaction (time:group)
# Random effects: Intercept grouped by 'id' to account for repeated measures within subjects
model.intervention <-  lmer(glucose_uptake ~ time * group +
                   (1 | id),
              data = mirage_data_long)

model.intervention

# Within group comparisons
emm_interaction <- emmeans(model.intervention, specs = pairwise ~ time | group)
pairwise <- pairs(emm_interaction)
conf_intervals <- confint(emm_interaction, level = 0.95) # For 95% confidence intervals
print(conf_intervals)
summary(pairwise)


# Between group comparisons
emm_betweengroup <- emmeans(model.intervention, specs = ~ time * group)
contrast_emm <- contrast(emm_betweengroup, interaction = "pairwise")
conf_intervals <- confint(contrast_emm, level = 0.95) # For 95% confidence intervals
print(conf_intervals)
summary(contrast_emm)



# Make function of the above with correction list () as an input.

# Spearman analysis function for lipidomics. Correction list.

# Make heatmap or so based on correlations.
```
